Altering Backward Pass Gradients to improve Convergence

### 2022-11-17 0100.tex
Mod3

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{wrapfig}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\raggedbottom
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

@@

\textbf{Type I:} Independent gradient modification at multiple points: In this method,
the gradients are first computed using standard procedure and then individually altered
as shown in Fig. \ref{fig:gradient_altering}(b). Gradient clipping
\cite{pascanu2013difficulty} and Adaptive gradient clipping \cite{brock2021high} are
examples of such modifications. In both of these methods, the gradients are first
computed using standard rules and then they are modified using some function. It can be
described as:

@@

\vspace{-0.5cm}
\begin{equation} \begin{split} p_i' \geq p_i & \text{,\ \ if } p_i \leq
\Big(\sum _{j=1}^C p_j^{\alpha }\Big)^{\frac{1}{\alpha-1}} \\ \hspace*{1cm}
p_i' < p_i & \text{,\ \ if } p_i > \Big(\sum _{j=1}^C
p_j^{\alpha}\Big)^{\frac{1}{\alpha-1}} \end{split} \label{eqn:threshold}
\end{equation}

@@

\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-4-2}
\caption{Layer 11\\ \textbf{(baseline)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-4-2}
\caption{Layer 11\\ \textbf{(PGT)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-7-2}
\caption{Final layer\\ \textbf{(baseline)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-7-2}
\caption{Final layer\\ \textbf{(PGT)}
}\end{subfigure}
\captionsetup{font=scriptsize}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-f-layer-22-1}
\caption{Norm of final layer's features\\ \textbf{(baseline)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-f-layer-22-1}
\caption{Norm of final layer's features\\ \textbf{(PGT)}
}\end{subfigure}

@@

\captionsetup{font=footnotesize}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-5-3}
\caption{Layer 15\\ \textbf{(baseline)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-5-3}
\caption{Layer 15\\ \textbf{(PGT)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-5-3}
\caption{Layer 15\\ \textbf{(AGC+PGT)}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-1-2}
\caption{Layer 2\\ \textbf{(baseline)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-1-2}
\caption{Layer 2\\ \textbf{(PGT)}
}\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-1-2}
\caption{Layer 2\\ \textbf{(AGC+PGT)}}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Norm vs iter. plots demonstrating layer characteristics (the zero out
phenomena) and the efficacy of PGT and AGC+PGT over baseline. Each colour represents a
different filter or feature vector of a particular layer of a non-BN variant of
ResNet-18. }
\label{fig:norm_plots}
\vspace{-0.5cm}
\end{figure}

Fig. \ref{fig:norm_plots}(c) and \ref{fig:norm_plots}(e) are the plot of the final conv
layer's filters and the final global average pooling layer respectively. We observe a
number of filters completely zeroing out. The feature vector here directly interfaces
with the fully connected layer immediately after. Therefore any zeroed out features
leads to permanent information loss, as it does not contribute to the learning of
decision boundaries in the fully connected layer. Also, zeroed out weights tensors lead
to zeroed out gradients hence stopping training for all subsequent iterations leading to
a collapse in training for the affected layers. With large batch sizes, it is possible
that an entire layer zeroes out [Fig. \ref{fig:high_bs}]. Gradient modification methods
such as PGT or AGC can alleviate the zeroing out phenomena. We observe that the number
of zeroed out filters has considerably reduced [Fig. \ref{fig:norm_plots}(b, d)]. The
final feature tensor [Fig. \ref{fig:norm_plots}(f)] with PGT enabled, does not contain
any zeroed out regions indicating that information loss is mitigated as the features
pass on from the feature extracting layers to the fully connected layer. When PGT and
AGC are combined, we get the most benefit for layer $15$ [Fig \ref{fig:norm_plots}(g-i)]
and layer $2$ [Fig \ref{fig:norm_plots}(j-l)].

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\columnwidth]{r18_grid_search}
\caption{ Grid search for various values of $\alpha$ for ResNet-18 architecture with the
ImageNet dataset. Each datapoint is collected over a 50 epoch budget, with other
hyperparameters kept the same as mentioned in section \ref{sec:bn}. }
\label{fig:grid_searcha}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\columnwidth]{r50_grid_search}
\caption{ Grid search for various values of $\alpha$ for ResNet-50 architecture with the
ImageNet dataset. Each datapoint is collected over a 50 epoch budget, with other
hyperparameters kept the same as mentioned in section \ref{sec:bn}. }
\label{fig:grid_searchb}
\end{figure}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{url}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{wrapfig}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\graphicspath{{images/}}
\raggedbottom
\newcommand{\forceindenta}{\parindent=1em\indent\parindent=0pt\relax}
\newcommand{\forceindentb}{\parindent=1.65em\indent\parindent=0pt\relax}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\NewDocumentCommand{\codeword}{v}{\texttt{\textcolor{orange}{#1}}}
\newcommand{\cmark}{\ding{51}} \newcommand{\xmark}{\ding{55}}
\definecolor{newcolor}{rgb}{.8,.349,.1}
\definecolor{codegreen}{rgb}{0.7,0.7,0.7}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{1,1,1}
\lstdefinestyle{mystyle}{ backgroundcolor=\color{backcolour},   
commentstyle=\color{codegreen}, keywordstyle=\color{magenta},
numberstyle=\tiny\color{codegray}, stringstyle=\color{codepurple},
basicstyle=\ttfamily\footnotesize, breakatwhitespace=false,         
breaklines=true,                 
captionpos=b,                    
keepspaces=true,                 
numbersep=5pt,                  
showspaces=false,                
showstringspaces=false, showtabs=false,                  
tabsize=2 }
\lstset{style=mystyle}

@@

\textbf{Type I:} Independent gradient modification at multiple points: Here the
gradients are first computed using standard procedure and then individually altered as
shown in Fig. \ref{fig:gradient_altering}(b). Gradient clipping
\cite{pascanu2013difficulty} and Adaptive gradient clipping \cite{brock2021high} are
examples of such modifications. In both of these methods, the gradients are first
computed using standard rules and then they are modified using some function. It can be
described as:

@@

\vspace{-0.5cm}
\begin{equation} \begin{split} p_i' \geq p_i & \text{,\ \ if } p_i \leq
\Big(\sum _{j=1}^C p_j^{\alpha }\Big)^{\frac{1}{\alpha-1}} \\ \hspace*{1cm}
p_i' < p_i & \text{,\ \ if } p_i > \Big(\sum _{j=1}^C
p_j^{\alpha}\Big)^{\frac{1}{\alpha-1}} \end{split} \label{eqn:threshold0}
\end{equation}

@@

\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-4-2}
\caption{Layer 11\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-4-2}
\caption{Layer 11\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-7-2}
\caption{Layer 19\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-7-2}
\caption{Layer 19\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-f-layer-22-1}
\caption{\scriptsize{GAP\\\forceindentb features}\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-f-layer-22-1}
\caption{\scriptsize{GAP\\\forceindentb features}\\ \forceindentb\textbf{PGT}}
\end{subfigure}

@@

\captionsetup{font=footnotesize}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Norm vs iter. plots demonstrating layer characteristics (the zero out
phenomena) and the efficacy of PGT and AGC+PGT over baseline. Each colour represents a
different filter or feature vector of a particular layer of a non-BN variant of
ResNet-18. }
\label{fig:norm_plots}
\vspace{-0.5cm}
\end{figure}

Fig. \ref{fig:norm_plots}(c) and \ref{fig:norm_plots}(e) are the plot of the final conv
layer's filters (layer $19$) and the features output of global average pooling (GAP)
layer respectively. We observe a number of filters in the final layer completely zeroing
out. The feature vector after GAP layer directly interfaces with the fully connected
layer immediately after. Therefore any zeroed out features leads to permanent
information loss, as it does not contribute to the learning of decision boundaries in
the fully connected layer. Also, zeroed out weights tensors lead to zeroed out gradients
hence stopping training for all subsequent iterations leading to a collapse in training
for the affected layers. With large batch sizes, it is possible that an entire layer
zeroes out [Fig. \ref{fig:high_bs}]. Gradient modification methods such as PGT or AGC
can alleviate the zeroing out phenomena. We observe that the number of zeroed out
filters has considerably reduced [Fig. \ref{fig:norm_plots}(b, d)]. The final feature
tensor [Fig. \ref{fig:norm_plots}(f)] with PGT enabled, does not contain any zeroed out
regions indicating that information loss is mitigated as the features pass on from the
feature extracting layers to the fully connected layer. When PGT and AGC are combined,
we get the most benefit for layer $15$ [Fig \ref{fig:norm_plots}(g-i)] and layer $2$
[Fig \ref{fig:norm_plots}(j-l)].

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\columnwidth]{r18_grid_search}
\caption{ Grid search for various values of $\alpha$ for ResNet-18 architecture with the
ImageNet dataset. Each datapoint is collected over a 50 epoch budget, with other
hyperparameters kept the same as mentioned in section \ref{sec:Expe}. }
\label{fig:grid_searcha}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\columnwidth]{r50_grid_search}
\caption{ Grid search for various values of $\alpha$ for ResNet-50 architecture with the
ImageNet dataset. Each datapoint is collected over a 50 epoch budget, with other
hyperparameters kept the same as mentioned in section \ref{sec:Expe}. }
\label{fig:grid_searchb}
\end{figure}
